---
title: "Spatial_Intro"
author: "Meggi Varnai"
date: "2023-01-25"
output: html_document
---
#Spatial Intro
```{r setup, include=FALSE}
#bringing our setup file to load packages
source("setup.R")
```

##Tigris vector data 
The tigris package is one of many data retrieval R packages that uses API calls to pull in data from various online/open databases directly into your R session

```{r import counties}
#import using tigris package, returns a shapefile consisting of polygons for each county
counties<- counties(state= "CO")
```


```{r}
# importing roads for Larimer county, which returns a polyline dataset for all roads in Larimer County
roads <- roads(state = "CO", county = "Larimer")
```


## quick visualization with tmap
```{r}
#interactive(us view) or static (use plot)
tmap_mode("view")
#qtm = quick thematic map 
qtm(counties)+
  qtm(roads)
#more detailed map, better for customization
tm_shape(counties)+
  tm_polygons() #telling it what kind of layer
```

looking at the structure/data types of our data
```{r}
class(counties)
#sf, data.frame
```
sf - simple feature (also an r package) helps work with vector data lets you work on spatial data like a 'normal' dataframe


##Point data
```{r point data}
#from lesson plan
poudre_points <- data.frame(name = c("Mishawaka", "Rustic", "Blue Lake Trailhead"),
                            long = c(-105.35634, -105.58159, -105.85563),
                            lat = c(40.68752, 40.69687, 40.57960))
```

### making non-spatial data spatial with the sf package
first, nonspatial object, telling the package our points using 'coords' and then the coordinate reference system (WGS-84 = 4326 (most of USA))
```{r}
poudre_points_sf <- st_as_sf(poudre_points, coords= c("long","lat"), crs=4326)
```

## raster data
elevatr package, returns a DEM
```{r}
#if you give it spatial, z= zoom level, the raster resolution
elevation <- get_elev_raster(counties, z= 7) #7= 1 km resoution
#class is raster layer

```
###mapping example
```{r}
qtm(elevation)
#since we downloaded the data this way, it is a 'temp' file, only exists in our current session
#more detailed map 
tm_shape(elevation)+
  tm_raster(style="cont", title= "Elevation (m") 
#want continuous color pallet and a title
```

##terra
working with the raster data in this class
terra quirk: only deals with SpatRaster, need to 
```{r}
#rast is from terra, makes RastLayer into SpatRaster
elevation<- rast(elevation)
#renames the str version where name was a long weird file and now it'll be "Elevation"
names(elevation)<- "Elevation"
```

before cropping, check the projection of the data
```{r}
st_crs(counties)
#NAD83 
crs(counties)== crs(elevation)
#FALSE
#project elevation layer
elevation_prj<- project(elevation, counties)
#doesnt work, need to specify which package to use!
elevation_prj<- terra::project(elevation, counties)

```

cropping to our AOI
```{r}
elevation_crop<- crop(elevation, ext(counties))

qtm(elevation_crop)
```

##read/write spatial data
saving sf or raster data
```{r}
write_sf(counties, "data/counties.shp")

#for raster, geotiff, .tiff format, what we will be using
writeRaster(elevation_crop,"data/elevation.itff")

```

save rdata file
```{r}
save(counties, roads, file= "data/spatial_objects.RData")
```


```{r}

```

```{r}

```

# Exercises

Filter out the counties data set to only include Larimer, Denver, and Pueblo counties.
```{r}

```

Make a map of the counties data colored by county area. Make a second map of counties colored by their total area of water.
```{r}

```

Make a barplot comparing the elevation of your 3 points in the Poudre Canyon (note: explore the extract() function in the terra package).
```{r}

```

Why are there 4 features in our Poudre Canyon Highway variable instead of 1?
